---
title: "Expression Workshop"
author: "Humberto Ortiz-Zuazaga"
date: "November 14, 2014"
output: html_document
---

## Abstract

The bioconductor project publishes software and data for the analysis of functional genomics experiments, using a variety of techniques such as microarrays or second-generation sequencing. We will explore the bioconductor tools for expression analysis and pathway analysis and apply them to a demo dataset.

## Installing bioconductor

Bioconductor comes with it's own installation procedures for software and data packages, the `biocLite()` function. You can use it to install any bioconductor package, along with all prerequisites. For example, to install the `limma` package, we can use code like this:

```
source("http://bioconductor.org/biocLite.R")
biocLite("limma")
```

Once you have sourced the biocLite file in your R session, you can continue to use it to install additional packages, such as the pathview package:

```
biocLite("pathview")
```

## Installing Data packages

Bioconductor includes many experimental data packages as well, that can be installed with `biocLite()` function as well:

```
biocLite("breastCancerMAINZ")
biocLite("hgu133a.db")
```

## Loading packages

Once packages have been installed, you need to load the package into each session that will use them. R-Studio will keep the packages loaded into a workspace, even if you exit and restart the program, as long as you save your workspace.

```{r}
library(Biobase)
library(limma)
library(breastCancerMAINZ)
data(mainz)
```


## Getting help.

bioconductor has extensive help available for almost all aspects.

```{r}
?mainz
?ExpressionSet
?limma
```

Of particular note, are the package vignettes, pdf files with tutorial examples of almost all packages.

## Examining the mainz breast cancer experiment

http://cancerres.aacrjournals.org/content/68/13/5405.abstract

```{r}
dim(mainz)
head(pData(mainz))
pData(mainz)$e.dmfs
# count patients with distant metastasis before 5 years
sum(pData(mainz)$e.dmfs & pData(mainz)$t.dmfs < 1825)
head(featureNames(mainz))
annotation(mainz)
```


```{r}
library(hgu133a.db)
```
