---
title: "Using online expression data in Bioconductor"
author: "Humberto Ortiz-Zuazaga"
date: "4/30/2021"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.12")

if (!requireNamespace("GEOquery", quietly = TRUE))
    install.packages("GEOquery")

if (!requireNamespace("limma", quietly = TRUE))
    install.packages("limma")

if (!requireNamespace("recount3", quietly = TRUE))
    install.packages("recount3")
```

## Paper

Walsh LA, Alvarez MJ, Sabio EY, Reyngold M, Makarov V, Mukherjee S, Lee KW, Desrichard A, Turcan Ş, Dalin MG, Rajasekhar VK, Chen S, Vahdat LT, Califano A, Chan TA. An Integrated Systems Biology Approach Identifies TRIM25 as a Key Determinant of Breast Cancer Metastasis. Cell Rep. 2017 Aug 15;20(7):1623-1640. doi: 10.1016/j.celrep.2017.07.052. PMID: 28813674; PMCID: PMC5985663.

## Data!

At the very end of the paper it says:

The accession numbers for the RNA-seq after TRIM25 knockdown in breast cancer cell lines, ChIP-seq in TNBC cell lines, and RIP-seq in TNBC cell lines data reported in this paper are GEO: GSE79589, GSE79588, and GSE79587, respectively. The accession number for the primary breast cancer and matched metastases microarray data reported in this paper is GEO: GSE57968.

# Microarray data

## Gene Expression Omnibus

NCBI resource for gene expression data.

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE57968

## GEO2R

Many analisis can be done in browser selecting GEO2R

1. sort Sample table by Source Name
1. create primary and metastasis group
1. assign rows to each group
1. click analyze

You can see the R script used to perform the analysis, save it, and alter it locally.

## Preparing

```{r}
#   Differential expression analysis with limma
library(GEOquery)
library(limma)
#library(umap)
```

## Loading

```{r}
# load series and platform data from GEO

gset <- getGEO("GSE57968", GSEMatrix =TRUE, AnnotGPL=TRUE)
if (length(gset) > 1) idx <- grep("GPL571", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]
```

## Additional data

```{r}
# make proper column names to match toptable 
fvarLabels(gset) <- make.names(fvarLabels(gset))

# group membership for all samples
gsms <- paste0("01010101010101010101010101010101010101010101010101",
        "0101010101010101010101")
sml <- strsplit(gsms, split="")[[1]]
```

## Design

```{r}
# assign samples to groups and set up design matrix
gs <- factor(sml)
groups <- make.names(c("primary","metastasis"))
levels(gs) <- groups
gset$group <- gs
design <- model.matrix(~group + 0, gset)
colnames(design) <- levels(gs)
```

## Making the fit

```{r}
fit <- lmFit(gset, design)  # fit linear model
```

## Choosing the contrasts

```{r}
# set up contrasts of interest and recalculate model coefficients
cts <- paste(groups[1], groups[2], sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
```

## Top table

```{r}
# compute statistics and table of top significant genes
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=10)
# summarize test results as "up", "down" or "not expressed"
dT <- decideTests(fit2, adjust.method="fdr", p.value=0.05)

tT
```

## Volcanoplot

```{r}
# volcano plot (log P-value vs log fold change)
colnames(fit2) # list contrast names
ct <- 1        # choose contrast of interest
volcanoplot(fit2, coef=ct, main=colnames(fit2)[ct], pch=20,
  highlight=length(which(dT[,ct]!=0)), names=rep('+', nrow(fit2)))

```

## Raw Data

GEO also stores raw data for many experiments

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE57968

# RNASeq data

## Recount3

http://rna.recount.bio/

## Loading

See http://rna.recount.bio/docs/quick-access.html#quick-recount3

```{r}
## Load recount3 R package
library("recount3")
```

## Projects

```{r}
## Find all available human projects
human_projects <- available_projects()
```

## particular project

```{r}
## Find the project you are interested in,
## here we use SRP072302 as an example
proj_info <- subset(
    human_projects,
    project == "SRP072302" & project_type == "data_sources"
)
```

## Load the data

```{r}
## Create a RangedSummarizedExperiment (RSE) object at the gene level
rse_gene_SRP072302 <- create_rse(proj_info)
```

```{r}
## One row per sample, one column per phenotype variable
dim(colData(rse_gene_SRP072302))
```

```{r}
colnames(colData(rse_gene_SRP072302))
```


## Load SRA table

```{r}
SraRunTable <- read.csv("SraRunTable.txt")
```

## Sort SRA table

```{r}
rse_gene <- rse_gene_SRP072302
sra <- SraRunTable

## Re-organize the SRA table based on the SRA Run IDs we have
sra <- sra[match(colData(rse_gene)$external_id, sra$Run), ]

## Append the variables of interest
sra_vars <- c("Strain")
colData(rse_gene) <- cbind(colData(rse_gene), strain = sra[, sra_vars])
```

## deseq2

```{r}
library(DESeq2)
```

```{r}
dds <- DESeqDataSet(rse_gene, design = ~ strain)
```


# Raw RNASeq data

## Finding the data

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79589

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79586

https://www.ncbi.nlm.nih.gov/sra?term=SRP072302

## SRA Tools

https://trace.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=toolkit_doc

On boqueron:

```
$ module load sratoolkit/2.10.4
```

## Some interesting Runs

From the SraRunTable.txt

SRR3290869
SRR3290870
SRR3290871
SRR3290872
SRR3290873
SRR3290874

## fastq-dump

```
$ fastq-dump -X 3 -Z SRR3290870
Read 3 spots for SRR3290870
Written 3 spots for SRR3290870
@SRR3290870.1 1 length=100
NATGAATTAAGTCATGGTACACATCTGCAAAATCTTCATCCCAGCTGGGTCTCTGACTCAGTTTATAGAAGAAGAATCCCTCAGCGATTATGATAGAGAT
+SRR3290870.1 1 length=100
#33<AEBCGGDBGGEGC=E@GGCGFE>1BEGGFGEGGG>1C<C>FDGGG:CBCBB>GGGGGGCGCGDFCGGGGG1GD@CFCGGGGCEGGGGGGFGGDD>B
@SRR3290870.2 2 length=100
NGGAGTTTGAGACCAGCCTGGCCAACATGGTGAAACCTCATCTCTACTAAGACAAAGTCTTGCTCTTGTTCCCCAGGCTGAAGTGCAATGGCACAATCTC
+SRR3290870.2 2 length=100
#<<ABGGGGGGGGGGGGGGGGGGGGGGGGGBGGGGGGGGGGGGGGGGGGGBCCC@GGGGGGGGGGGGGGGCDGGGGGGGGDGGGGGGGGGGGGGGGGGGG
@SRR3290870.3 3 length=100
NCGCAGGAAGCCTCCACCCCCACCCCTGCCTGGCAGGTCCCAGATTACATGGGTGGCAGCTTCGGGGACAATCTGGTCACCCGCTCCTACCTCCTGGGCA
+SRR3290870.3 3 length=100
#<3<AECFF1CFGDGEGFGGGGGGGGGGGGG>GGE<GGGGGGGEGGGGGEB:?BBGEFG<F@GGDCGGAB0:CC@1EGFC>FGGG>9GEFGGGGGCFEGG
```

## Automating

```
$ for file in $(cat runs.txt) ; do fastq-dump -X 10 --split-files $file ; done
Read 10 spots for SRR3290869
Written 10 spots for SRR3290869
Read 10 spots for SRR3290870
Written 10 spots for SRR3290870
2021-04-30T17:22:50 fastq-dump.2.10.4 err: timeout exhausted while creating file within network system module - error with https open 'https://sra-pub-run-odp.s3.amazonaws.com/sra/SRR3290871/SRR3290871'
2021-04-30T17:22:50 fastq-dump.2.10.4 err: timeout exhausted while creating file within network system module - failed to open 'SRR3290871'
fastq-dump (PID 28391) quit with error code 3
```

